---
title: "Bicycle Rentals exploration by Sreejith Gopinath"
output:
  html_document:
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
    theme: united
    highlight: tango
    fig_caption: true
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```
# Univariate plots section

```{r, echo=FALSE}
setwd('/Users/sreeg/Exploratory_data_analysis')
df <- read.csv('hour.csv')
df_daily <- read.csv('day.csv')
df_master <- read.csv('hour.csv')

hr_casreg <- subset(df, select=c("hr","casual","registered"))
df <- subset(df, select = -c(casual, registered))
```

<span style="color:black">These are the dimensions of the bicycle_rentals 
dataset</span>

```{r, echo=FALSE}
dim(df)
```

<span style="color:black">The features in this dataset are:</span>
```{r, echo=FALSE}
colnames(df)

#Change values of certain columns from mere numbers to meaningful words
df$season[df$season == 1] <- "spring"
df$season[df$season == 2] <- "summer"
df$season[df$season == 3] <- "fall"
df$season[df$season == 4] <- "winter"

df$weathersit[df$weathersit == 1] <- "clear"
df$weathersit[df$weathersit == 2] <- "mild"
df$weathersit[df$weathersit == 3] <- "light precip"
df$weathersit[df$weathersit == 4] <- "heavy precip"

df$weekday[df$weekday == 0] <- "Sunday"
df$weekday[df$weekday == 1] <- "Monday"
df$weekday[df$weekday == 2] <- "Tuesday"
df$weekday[df$weekday == 3] <- "Wednesday"
df$weekday[df$weekday == 4] <- "Thursday"
df$weekday[df$weekday == 5] <- "Friday"
df$weekday[df$weekday == 6] <- "Saturday"

df_daily$weekday[df_daily$weekday == 0] <- "Sunday"
df_daily$weekday[df_daily$weekday == 1] <- "Monday"
df_daily$weekday[df_daily$weekday == 2] <- "Tuesday"
df_daily$weekday[df_daily$weekday == 3] <- "Wednesday"
df_daily$weekday[df_daily$weekday == 4] <- "Thursday"
df_daily$weekday[df_daily$weekday == 5] <- "Friday"
df_daily$weekday[df_daily$weekday == 6] <- "Saturday"

df_master$weekday[df_master$weekday == 0] <- "Sunday"
df_master$weekday[df_master$weekday == 1] <- "Monday"
df_master$weekday[df_master$weekday == 2] <- "Tuesday"
df_master$weekday[df_master$weekday == 3] <- "Wednesday"
df_master$weekday[df_master$weekday == 4] <- "Thursday"
df_master$weekday[df_master$weekday == 5] <- "Friday"
df_master$weekday[df_master$weekday == 6] <- "Saturday"

df$DoW <- factor(df$weekday, levels = c("Sunday", "Monday", "Tuesday", 
                                       "Wednesday", "Thursday", "Friday", 
                                       "Saturday"))
df_daily$DoW <- factor(df_daily$weekday, levels = c("Sunday", "Monday", 
                                                    "Tuesday", "Wednesday", 
                                                    "Thursday", "Friday", 
                                                    "Saturday"))
df_master$DoW <- factor(df_master$weekday, levels= c("Sunday", "Monday", 
                                                     "Tuesday", "Wednesday", 
                                                     "Thursday", "Friday", 
                                                     "Saturday"))
#Print out first few records
head(df)

#Print out unique values of a few columns
unique(df$weathersit)
unique(df$season)

str(df)

summary(df)
```
<span style="color:black"> <br>Here is a complete list of features in the 
dataset, along with their meanings:<br><br></span>
<table border="1">
  <tr>
    <th>Feature</th>
    <th>Explanation</th>
  </tr>
  <tr>
    <th>instant</th>
    <td><i>Unique identifier for each observation in the dataset.</i></td>
  </tr>
  <tr>
    <th>dteday</th>
    <td><i>The date for which the observation was recorded.</i></td>
  </tr>
  <tr>
    <th>season</th>
    <td><i>The prevailing season for which the observation was recorded.
  <br>1: Spring<br>2: Summer<br>3: Fall<br>4: Winter</i></td>
  </tr>
  <tr>
    <th>yr</th>
    <td><i>The year for which the observation was recorded. 0: 2011, 1:2012.
    </i><br></td>
  </tr>
  <tr>
    <th>mnth</th>
    <td><i> The month for which the observation was recorded. 1-Jan to 
12-Dec.</i></td>
  </tr>
  <tr>
    <th>hr</th>
    <td><i> The hour, in 24-hour-format, for which the observation was
recorded.</i></td>
  </tr>
  <tr>
    <th>holiday</th>
    <td><i>Whether the observation was recorded for holiday or a working day.
    </i></td>
  </tr>
  <tr>
    <th>weekday</th>
    <td><i>The day of the week for which the observation was recorded.</i><br>
    </td>
  </tr>
  <tr>
    <th>workingday</th>
    <td><i> If the day of observation is neither weekend nor holiday, then this
    value is 1, otherwise it is 0.</i></td>
  </tr>
  <tr>
    <th>weathersit</th>
    <td><i>Weather situation prevailing on the day of observation:
      <br> <t>- 1: Clear, Few clouds, Partly cloudy, Partly cloudy<br>
           <t>- 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, 
           Mist<br>
           <t>- 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, 
           Light Rain + Scattered clouds<br>
           <t>- 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
           </i><br></td>
  </tr>
  <tr>
    <th>temp</b></th>
    <td><i>Normalized temperature in Celsius, on the day of observation. The 
           values are divided to 41 (max).</i></td>
  </tr>
  <tr>
    <th>atemp</b></th>
    <td><i>Normalized feeling temperature in Celsius. The values are divided to 
    50 (max).</i></td>
  </tr>
  <tr>
    <th>hum</th>
    <td><i>Normalized humidity. The values are divided to 100 (max).</i></td>
  </tr>
  <tr>
    <th>windspeed</th>
    <td><i>Normalized wind speed. The values are divided to 67 (max).</i></td>
  </tr>
  <tr>
    <th>casual</th>
    <td><i>count of casual users.</i></td>
  </tr>
  <tr>
    <th>registered</th>
    <td><i> count of registered users.</i></td>
  </tr>
  <tr>
    <th>cnt</th>
    <td><i>count of total rental bikes including both casual and registered.</i>
    </td>
  </td>
</table>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

## <i>Sounds interesting!!</i><br>
<span style="color:black"><br>We see that the minimum ridership values on any 
given day is 1 and the maximum on any given day is just shy of 1000.</span>

```{r, echo=FALSE}
lowest_ridership <- subset(df_daily[order(df_daily$cnt, decreasing=F),])
lowest_ridership <- lowest_ridership[,c("dteday","cnt")]
head(lowest_ridership)
```

<span style="color:black">We see from the preceding dataframe dump that the 
lowest ever bicycle rentals in our dataset happened on 29th October 2012. A 
simple web search shows us that this was the day Hurricane Sandy made landfall 
in the Washington DC area - a majority of schools, colleges and offices had 
shut down! The next low occurred on 27th January 2011. On this day, it was a 
blizzard that had shut down Washington DC! Thus, analyzing transportation 
statistics over a period of time and co-relating incidents that correspond to 
variations to the trend, may help us build predictive models.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

<span style="color:black">Let us look at some plots now.<br></span>
```{r, echo=FALSE}
library(ggplot2)
library(reshape2)
library(GGally)
library(scales)
library(memisc)
library(lattice)
library(MASS)
library(plyr)
```{r, out.width = '33%', echo = FALSE, message = FALSE}
ggplot(aes(x=yr, y=cnt), data=df_master) + 
  geom_bar(fill="#006633", stat="identity") + 
  xlab("Year") +
  ylab("Total count") +
  ggtitle("Total users year-over-year")
ggplot(aes(x=yr, y=casual), data=df_master) + 
  geom_bar(fill="#006633", stat="identity") + 
  xlab("Year") +
  ylab("Casual user count") +
  ggtitle("Casual users year-over-year")
ggplot(aes(x=yr, y=registered), data=df_master) + 
  geom_bar(fill="#006633", stat="identity") + 
  xlab("Year") +
  ylab("Registered user count") +
  ggtitle("Registered users year-over-year")
```

<span style="color:black">The first plot on the left shows the total number of 
users renting bicycles in each of the two years for which we have data. At a 
very high level, we see that the total number of users in the initial year are 
outnumbered by the total number of users in the subsequent year. I attribute 
this trend to a number of possible factors like rise in popularity of the rental 
service, increase in the travelling population and increase in the crowding 
factor on other modes of public transportation.<br><br>The second plot shows the 
total number of casual users renting bicycles in the two years for which we have 
data. We see that the total number of casual users in the initial year are 
outnumbered by the total number of users in the subsequent year.<br><br>The last 
plot shows the total number of registered users renting bicycles in the two 
years for which we have data. We see that the total number of registered users 
in the initial year are outnumbered by the total number of users in the 
subsequent year. We also see that the year-over-year increase in registered 
users is greater than the year-over-year increase in casual users. This may be 
due to casual-to-registered user conversion, or simply more people choosing to 
register than rent casually.<br><br>I chose to include these plot because I
wanted to look at the overall trends and whether there were any opposing trends 
among casual and registered users. We do not see any opposing trends, but it 
looks like registered ridership is signaling the popularity of the service.
</span>

```{r, echo=FALSE}
ggplot(aes(cnt), data=df) + 
  geom_histogram(binwidth=10, position="identity", fill="#006633") + 
  scale_y_log10() + 
  xlab("Year") +
  ylab("Frequency") +
  ggtitle("Total bicycle rentals")
```

<span style="color:black">The bicycle rental frequencies seem to progressively 
decline for higher counts. I chose to include this plot because I wanted to 
convey a sense of popularity for the service. We see that daily/hourly instances
with lower total bicycle rentals happen more frequently than daily/hourly 
instances with higher total rentals.<br><br>I wonder how the frequencies will 
vary by season and weather conditions.</span>

```{r, echo=FALSE}
ggplot(aes(x=weathersit, y=cnt), data=df) + 
  stat_summary(fun.y=median, geom="bar", fill="#006633") + 
  xlab("Prevailing weather situation") +
  ylab("Median rental count") +
  ggtitle("Median bicycle rentals by prevailing weather situation")
```

<span style="color:black">As is intuitive, we see that bicycle rentals are 
highest when the weather is clear, next highest when the weather is just misty, 
a few brave souls renting bicycles when there is light precipitation and least 
when there is heavy precipitation. I chose to include this plot because bicycle
transportation is open to the elements and is therefore very different from 
other modes of transportation. I wanted to get an initial sense of whether the
prevailing conditions played a major role in influencing bicycle rentals.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=cnt), data=df) + 
  stat_summary(fun.y=median, geom="bar", fill="#006633") + 
  xlab("Prevailing season") +
  ylab("Median rental count") +
  ggtitle("Median hourly total bicycle rentals by season")
```

<span style="color:black">I cannot fathom the reason for this, but it looks like
bicycle rentals are highest in the fall, next highest in summer, third in winter
and lowest in spring. As in the previous plot, I wanted to explore whether
prevailing conditions played a part in influencing bicycle rentals. Looking at
the seasonal trends, we get a fair sense of how general cold weather/warm 
weather, etc. influence bicycle rentals.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

<span style="color:black">Next, I subset-ed the data into two groups - one 
subset for holidays and one subset for working days. Here are the summaries of 
the bicycle rental numbers for the respective subsets:</span>

```{r, echo=FALSE}
df_working <- subset(df, holiday == 0)
df_holiday <- subset(df, holiday == 1)
summary(df_working$cnt)
summary(df_holiday$cnt)
```

<span style="color:black">From the preceding stats, we see that, all else set 
aside, bicycle rental numbers are higher on working days than on holidays - 
indicating that bicycle rentals may not be for recreational use.</span>

<span style="color:black">Next, let's explore how bicycle rentals stack up by 
day of the week.</span>

```{r,  out.width = '50%', echo = FALSE, message = FALSE}
ggplot(aes(x=DoW, y=cnt, group="DoW"), data=df) + 
  stat_summary(fun.y=mean, geom="line", colour="red") +
  xlab("Day of the week") +
  ylab("Mean daily count") +
  ggtitle("Mean daily total bicycle rentals")

ggplot(aes(x=hr, y=cnt), data=df) + 
  scale_x_continuous(breaks=-1:24) + 
  stat_summary(fun.y=median, geom="line", color="red") + 
  xlab("Hour of the day") +
  ylab("Median hourly count") +
  ggtitle("Median hourly total bicycle rentals")
```

<span style="color:black">From the plot on the left, we see that the busiest day 
for bicycle rentals is Friday, and the least busy day is Sunday. Rental numbers 
climb up through the work week and then fall off on the weekend.<br><br>From the
plot on the right, we see that bicycle rentals peak at 8:00am and 5:00pm and 
drop off on either side, suggesting that the peaks may be driven up by people 
using bicycles to get to and from work.<br><br>I chose to include these plots 
because I wanted to see whether there were any particular days that were busiest
and any that were less busy, as also the busier hours of the day as well the 
less busy ones.</span>

```{r,  out.width = '50%', echo = FALSE, message = FALSE}
ggplot(aes(x=DoW, y=cnt), data=df) + 
  geom_boxplot(fill="maroon") + 
  ylim(0, 300) +
  xlab("Day of the week") +
  ylab("Total rental count") +
  ggtitle("Total bicycle rentals by day of the week")
ggplot(aes(x=factor(hr), y=cnt), data=df) + 
  geom_boxplot(fill="maroon") + 
  ylim(0, 300) +
  xlab("Hour of the day") +
  ylab("Total rental count") +
  ggtitle("Total bicycle rentals by hour of the day")
```

<span style="color:black"><br>The plot on the left shows the daily total bicycle
rentals by the day of the week. In the previous section, we saw which were the
busiest days of the week for bicycle rentals; I chose this plot so that we could
see how noisy the data really is. During the weekend, we see the least variance,
while we see greater variance for data on Monday and Wednesday.<br><br>The plot 
on the right shows the hourly total bicycle rentals. In the previous section, we 
saw which were the busiest hours of the day for bicycle rentals; I chose this 
plot so that we could see how noisy the datareally is. This plot turned out to 
be really interesting; we see the medians following the trend we saw in the 
previous section and we also see that most of the variance in bicycle rental 
hours happens during the middle portion of the day. I have limited the y-axis to
a value of 300, so that the outliers are eliminated and the charts are most 
readable.</span>

```{r, echo = FALSE}
ggplot(aes(x=hr, y=cnt, col=factor(holiday)), data=df) + 
  scale_x_continuous(breaks=-1:24) + 
  stat_summary(fun.y=median, geom="line") + 
  xlab("Hour of day") +
  ylab("Median rental count") +
  ggtitle("Median hourly total bicycle rentals on working days")
```

<span style="color:black">The preceding plot shows the median total bicycle 
rentals on working days for each hour of the day(line in red). On working days, 
the trend is pretty close to the trends overall, suggesting that bicycle rentals 
are primarily driven by office/school-goers. This plot continues the exploration 
from where we left off in the previous plot. The line in blue shows 
bicycle rentals by the hour on holidays. The plot for holidays shows higher 
numbers during the middle portion of the day. I chose these plots because I 
wanted to see if the hourly trends on working days matched the hourly trends on 
all days. It does follow closely, reinforcing our belief that bicycle rentals 
are primarily used by commuters.</span>

<span style="color:black">Let us now see how the numbers of casual and 
registered renters vary by the day of the week and the hour of the day.</span>

```{r,  out.width = '50%', echo = FALSE, message = FALSE}
day_casreg <- subset(df_daily, select=c("DoW","casual","registered"))
day_casreg <- melt(day_casreg, id.vars="DoW")
ggplot(aes(x=DoW, y=value, fill=variable), data=day_casreg) + 
  geom_bar(position="dodge", binwidth=1, fun.y=median, stat="identity") +
  xlab("Day of the week") +
  ylab("Median daily rental count") +
  ggtitle("Median daily bicycle rentals")

hr_casreg <- melt(hr_casreg, id.vars="hr")
ggplot(hr_casreg, aes(hr, value, col=variable)) + 
  scale_x_continuous(breaks=-1:24) + 
  stat_summary(fun.y=median, geom="line") + 
  xlab("Hour of the day") +
  ylab("Median rental count") +
  ggtitle("Median hourly bicycle rentals")
```

<span style="color:black">From plot on the left, we see that casual 
ridership is higher on weekends than on weekdays, while the trend is reversed 
for registered ridership. We also see that the hourly ridership plot for 
registered riders is similar to the overall hourly ridership plot - which means 
that rental registration is a pretty good predictor of bicycle rentals.<br><br>
Though this plot may have better belonged in the bivariate section, I chose to
include this plot here because it nicely wraps up the conclusions drawn in the
preceding graphs. We see how registered users(more likely commuter than not)
drive up rentals in the middle portion of the week while casual(more likely to
be recreational users than not) drive up rentals during the weekend.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

# Univariate analysis

<span style="color:black"><b><i>What is the structure of your dataset?</i></b>
</span><br>
<span style="color:red">I am working with the bicycle rentals dataset, which 
contains bicycle rentals information at an agency in Washington DC., for the 
years 2011-2012. The dataset also contains weather data for the same days as the
bicycle rentals.<br><br>The dataset contains 731 daily records as well as 17379
hourly records for bicycle rentals at one agency in the city.<br><br>There are 
17 features in the hourly dataset, and they are listed here:<br><b>instant</b> -
An index for the record<br><b>dteday</b> - The date<br><b>season</b> - The 
prevailing climatic season<br><b>yr</b> - Year<br><b>mnth</b> - Month<br><b>hr
</b> - The hour of observation<br><b>holiday</b> - Whether or not the particular
day was a holiday(weekends are counted as holidays too)<br><b>weekday</b> - The 
day of the week<br><b>workingday</b> - Whether or not the particular day was a 
working day(inverse of the holiday value)<br><b>weathersit</b> - What kind of 
weather prevailed on the particular day<br><b>temp</b> - The hourly temperature
<br><b>atemp</b> - The average temperage for the day<br><b>hum</b> - The 
humidity on the particular day<br><b>windspeed</b> - The windspeed on the 
particular day<br><b>casual</b> - The number of rentals by non-registered 
customers<br><b>registered</b> - The number of rentals by registered customers
<br><b>cnt</b> - Total number of rentals</span>

<span style="color:red">The variables <i>weathersit</i>, <i>season</i>, 
<i>holiday</i> and <i>workingday</i> are categorical variables.<br><b>weathersit
</b> has four states that I have simplified into: <i>clear</i>, <i>misty</i>, 
<i>light_precipi</i>(light rain/snow) and <i>heavy_precipi</i>(heavy rain/sleet/
snow).<br><b>season</b> has four states: <i>spring</i>, <i>summer</i>, <i>fall
</i> and <i>winter</i>.<br><b>holiday</b> and <b>workingday</b> are categorical 
variables that have binary values and orthogonal with each other in each 
observation.<br><br>Other observations:<br>1) Most bicycle rentals are made by 
registered customers than casual customers.<br>2) More bicycle rentals are made 
by people that are commuting to their jobs/schools than recreational customers.
<br>3)Bicycle rentals are higher in dry weather than wet weather.</span>

<span style="color:black"><b><i>What is/are the main feature(s) of interest in 
your dataset?</i></b></span><br>
<span style="color:red">The main feature of interest in the dataset is the 
bicycle rental count. I am trying to build a prediction model for bicycle 
rentals. From our univariate analyses, we have seen that the general weather 
situation and season have pretty good predictive power.</span>

<span style="color:black"><b><i>What other features in the dataset do you think 
will help support your investigation into your feature(s) of interest?</i></b>
</span><br>
<span style="color:red">We need to analyse some more to build a prediction model
with higher confidence.<br>I suspect windspeed, humidity and temperature will 
play a role in predicting bicycle rentals, but we shall see.</span>

<span style="color:black"><b><i>Did you create any new variables from existing 
variables in the dataset?</i></b></span><br>
<span style="color:red">No, I did not. Had the total count of bicycle rentals 
been missing, I could have used "casual" and "registered" to derive the total 
count.</span>

<span style="color:black"><b><i>Of the features you investigated, were there any
unusual distributions? Did you perform any operations on the data to tidy, 
adjust, or change the form of the data? If so, why did you do this?</span></i>
</b><br>
<span style="color:red">I log-transformed the total bicycle rentals for each 
day, to get a sense for the distribution.<br>1) When I looked at the 
distribution of total bicycle rentals as a whole, I found that the lower numbers
occurred far more frequently. There was a tiny gap in the frequencies around the
mark, which remains unexplained.<br>2) When I plotted the busiest hours of the 
day overall, I saw two peaks, which corresponded to the commute hours.<br>3) 
When I looked at the mean trends on a daily basis, I see that registered users 
rented mainly during the work week and lesser on weekends. For casual users, 
this trend was reversed - casual users rented less during the work week than on 
weekends.<br><br>I did not have to do anything to clean the dataset.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

# Bivariate plots section

<span style="color:black">From the univariate analyses performed so far, I have 
seen that some features tend to correlate with each other. For example, average 
temperatures and precipitation seem to correlate with the season in a broader 
sense; these variables also seem to correlate with the weather situation in a 
much more specific sense.</span>

```{r, out.width='100%', echo=FALSE}
set.seed(2050)
df_sample <- df[sample(1:length(df$cnt), 5000),]
colnames(df_sample)[7] <- "hday"
colnames(df_sample)[8] <- "wday"
colnames(df_sample)[10] <- "wthr"
colnames(df_sample)[14] <- "wind"
options(warn=-1)
ggpairs(df_sample, columns=c(3,4,5,6,7,8,10,12,13,14,15), 
        params = c(shape = I('.'), outliers.shape = I('.'), list(corSize=5)), 
        axisLabels = "none")
```
<span style="color:black"><br><br><br>A quick look at the correlation data shows
that bicycle rentals are mostly influenced by the feeling temperature and hour 
of the day. I chose to include the correlation chart because it shows
correlation scores for different pairs of features and therefore will give us a
good idea of what relationships we should be exploring in detail.<br>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

<br>I will now explore some scatter plots involving total bicyle rentals 
together with normalized daily feeling temperatures, humidity and windspeed.
</span>

```{r, echo=FALSE}
ggplot(df, aes(x=atemp, y=cnt)) + 
  geom_jitter(alpha = 0.1, size=1) + 
  geom_smooth(method = "loess") + 
  xlab("Normalized feeling temperature") +
  ylab("Total rental count") +
  ggtitle("Scatterplot of total bike rentals Vs normalized feeling temperature")
```
<span style="color:black"><br><br>From the preceding plot, we can see that 
bicycle rentals gradually climbs up as the temperature increases, and is lowest 
at the extremeties, ie., when the feeling temperatures are at their lowest and 
highest. From the correlation chart above, we see a positive correlation between
the total bicycle rental numbers and the prevailing temperatures, with a 
correlation score of 0.397. I chose to include this plot because we saw in the 
previous section how prevailing weather conditions influence bicycle rentals. I 
wanted to see whether prevailing temperatures had any role to play.</span>

```{r, echo=FALSE}
ggplot(df, aes(x=hum, y=cnt)) +
  geom_jitter(alpha=0.1, size=1) + 
  geom_smooth(method = "loess") + 
  xlab("Normalized humidity") +
  ylab("Total rental count") +
  ggtitle("Scatterplot of total bicycle rentals Vs humidity")
```
<span style="color:black"><br><br>From the preceding plots, we see that most 
bicycle rentals occur when the normalized humidity is between 0.30-0.80.Taking a
closer look at the rental number distributions between normalized humidities of 
0.75-1.0, we see that there are certain "pockets" of distributions. At this 
point, it is not entirely clear what might be causing these distributions. It 
may be because certain values of normalized humidity never occur. From the 
correlation chart above, we see that there is a negative correlation
between the total bicycle rental numbers and the prevailing humidity, with a 
correlation score of -0.325. I chose to include this plot because it helps us 
reinforce in greater details the relationship between the total number of 
bicycle rentals and the prevailing humidity that was revealed by the correlation 
chart.</span>

```{r, echo=FALSE}
ggplot(df, aes(x=windspeed, y=cnt)) +
  geom_jitter(alpha = 0.1, size=1) + 
  geom_smooth(method = "loess") + 
  xlab("Windspeed") +
  ylab("Total rental count") +
  ggtitle("Scatterplot of total bicycle rentals Vs windspeed")
```

<span style="color:black"><br><br>From the preceding plots, we see that bicycle 
rentals are pretty steady at lower windspeeds, dropping off at higher windspeeds
. In the correlation chart above, we see that the correlation between total 
bicycle rental numbers and the prevailing windspeeds is posistive, but weak, 
with a correlation score of 0.0893. I chose to include this plot because it 
nicely wraps up our investigation into how the three main weather factors 
influence bicycle rentals individually.<br><br>I chose to include scatterplots 
in the above charts to investigate if there were any patterns. I did not see any 
particular patterns, other than the fact that a combination of moderate-high 
humidity, moderate temperature and low-moderate windspeeds might work well for 
bicycle renters!</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

```{r, echo=FALSE}
#Change values of certain columns from mere numbers to meaningful words
df$season[df$season == 1] <- "spring"
df$season[df$season == 2] <- "summer"
df$season[df$season == 3] <- "fall"
df$season[df$season == 4] <- "winter"

df$weathersit[df$weathersit == 1] <- "clear"
df$weathersit[df$weathersit == 2] <- "misty"
df$weathersit[df$weathersit == 3] <- "light_precipitation"
df$weathersit[df$weathersit == 4] <- "heavy_precipitation"

df$weekday[df$weekday == 0] <- "Sunday"
df$weekday[df$weekday == 1] <- "Monday"
df$weekday[df$weekday == 2] <- "Tuesday"
df$weekday[df$weekday == 3] <- "Wednesday"
df$weekday[df$weekday == 4] <- "Thursday"
df$weekday[df$weekday == 5] <- "Friday"
df$weekday[df$weekday == 6] <- "Saturday"
df$DoW <- factor(df$weekday, levels= c("Sunday", "Monday", "Tuesday", 
                                       "Wednesday", "Thursday", "Friday", 
                                       "Saturday"))
ggplot(aes(x=atemp, fill=weathersit), data=df) + 
  geom_bar(position="dodge", binwidth=0.05) +
  xlab("Normalized feeling temperature") +
  ylab("Total rental count") +
  ggtitle("Total bicycle rentals Vs normalized feeling temperatures")
```

<span style="color:black"><br><br>From the preceding plot, we see that bicycle 
rentals mostly happen in "moderate" feeling temperature. We see that clear 
weather rentals occur across the spectrum of feeling temperature. We also see 
that bicycle rentals in misty and light precipitation weather situations happen 
in a much narrower band of temperature. A peak stands out at a normalized 
feeling temperature of about 0.63, and we see that the rentals spike in clear, 
light precipitation and misty weather. Of course, very few rentals occur during 
periods of heavy precipitation.<br><br>Since the weather factors tend to 
straddle seasons and even occur unseasonally, I included this chart to explore 
how bicycle rentals vary with the feeling temperatures. Overlaying it with the
prevailing weather situation shows us the the different weather situations in
which the temperature values occur.</span>

```{r, echo=FALSE}
ggplot(aes(x=hum, fill=weathersit), data=df) + 
  geom_bar(position="dodge", binwidth=0.05) +
  xlab("Humidity") +
  ylab("Total rental count") +
  ggtitle("Total bicycle rentals Vs humidity")
```

<span style="color:black"><br><br>From the preceding plot, we see that most of 
the humidity range is dominated by clear weather and this is where a bulk of 
bicycle rentals take place. We do see a few peaks towards the higher end of the 
humidity spectrum, and this quartile seems to be dominated by misty weather for 
the most part and precipitative weather to a lesser extent.<br><br>Since the 
weather factors tend to straddle seasons and even occur unseasonally, I included 
this chart to explore how bicycle rentals vary with the prevailing humidity. 
Overlaying it with the prevailing weather situation shows us the the different 
weather situations in which the humidity values occur.</span>

```{r, echo=FALSE}
ggplot(aes(x=windspeed, fill=weathersit), data=df) + 
  geom_bar(position="dodge", binwidth=0.05) +
  xlab("Windspeed") +
  ylab("Total rental count") +
  ggtitle("Total bicycle rentals Vs windspeed")
```

<span style="color:black"><br><br>From the preceding plot, we see that most of 
the bicycle rentals happen in the first two quartiles of normalized windspeed. 
Rentals are highest when the normalized windspeed is 0, in fact. Beyond a 
normalized windspeed of 0.30, however, bicycle rentals fall away steadily and 
are almost non-exixtent in very windy weather. Most of the visible spectrum is 
dominated by clear weather.<br><br>Since the weather factors tend to straddle 
seasons and even occur unseasonally, I included this chart to explore how 
bicycle rentals vary with the prevailing windspeed. Overlaying it with the 
prevailing weather situation shows us the the different weather situations in 
which the windspeed values occur.<br><br>In fact, most of the bicycle rentals 
happen in clear weather, as shown by the summary below.</span>

```{r, echo=FALSE}
do.call(rbind , by(df$cnt, df$weathersit, summary))
```

<span style="color:black"><br>Let's also look at a summary of bicycle 
rentals by season.</span>

```{r, echo=FALSE}
do.call(rbind, by(df$cnt, df$season, summary))
```

<span style="color:black"><br>From the above summary, it looks like bicycle 
rentals are high in the fall and summer and taper off in the winter and spring.
<br><br>Let's now look at the number of casual bicycle rentals per season.
</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

```{r,  out.width = '50%', echo = FALSE, message = FALSE}
df_master$season[df_master$season == 1] <- "spring"
df_master$season[df_master$season == 2] <- "summer"
df_master$season[df_master$season == 3] <- "fall"
df_master$season[df_master$season == 4] <- "winter"
df_master$weathersit[df_master$weathersit == 1] <- "clear"
df_master$weathersit[df_master$weathersit == 2] <- "mild"
df_master$weathersit[df_master$weathersit == 3] <- "light precip"
df_master$weathersit[df_master$weathersit == 4] <- "heavy precip"
df_master$weekday[df_master$weekday == 0] <- "Sunday"
df_master$weekday[df_master$weekday == 1] <- "Monday"
df_master$weekday[df_master$weekday == 2] <- "Tuesday"
df_master$weekday[df_master$weekday == 3] <- "Wednesday"
df_master$weekday[df_master$weekday == 4] <- "Thursday"
df_master$weekday[df_master$weekday == 5] <- "Friday"
df_master$weekday[df_master$weekday == 6] <- "Saturday"

df_master$DoW <- factor(df_master$weekday, levels= c("Sunday", "Monday", 
                                                     "Tuesday", "Wednesday", 
                                                     "Thursday", "Friday", 
                                                     "Saturday"))

ggplot(aes(x=season, y=casual), data=df_master) + 
  stat_summary(fun.y=mean, geom="bar", fill="#006633") + 
  xlab("Prevailing season") +
  ylab("Casual rental count") +
  ggtitle("Casual bicycle rentals by prevailing season")
ggplot(aes(x=season, y=registered), data=df_master) + 
  stat_summary(fun.y=mean, geom="bar", fill="#006633") + 
  xlab("Prevailing season") +
  ylab("Mean rental count") +
  ggtitle("Registered bicycle rentals by prevailing season")
```

<span style="color:black">In the plot on the left, we see bar plots for casual
bicycle rentals by the prevailing season. We see that casual ridership is
highest in fall, followed by summer, spring and winter. In the plot on the 
right, we see bar plots for registered bicycle rentals by the prevailing season.
As in the case of casual users, we see that ridership is highest in summer. For 
registered users, though, next highest is in winter followed by spring. As in 
the case of casual users, lowest ridership occurs in spring.<br><br>I chose to 
include these plots because we can, at a glance, see how the seasonal trends
for bicycle rentals differs for casual renters and registered renters.</span>

```{r,  out.width = '50%', echo = FALSE, message = FALSE}
ggplot(aes(x=atemp, y=casual), data=df_master) + 
  geom_bar(fill="#006633", stat="identity") + 
  xlab("Normalized feeling temperatures") +
  ylab("Casual rental count") +
  ggtitle("Casual rentals Vs prevailing temperature")
ggplot(aes(x=atemp, y=registered), data=df_master) + 
  geom_bar(fill="#006633", stat="identity") + 
  xlab("Normalized feeling temperatures") +
  ylab("Registered rental count") +
  ggtitle("Registered bicycle rentals Vs prevailing temperature")
```

<span style="color:black">In the plot on the left, we see how casual bicycle
rentals stack up against the prevailing temperature. The highest number of 
casual bicycle rentals happen between normalized temperatures of 0.55 and 0.75.
<br><br>In the plot on the right, we see how registered bicycle rentals stack up 
against the prevailing temperatures. As in the case of casual riders, normalized
temperatures 0.48 to 0.75 seem to be most popular.<br><br>I chose to include 
these plots because it is interesting to see how the distribution of casual and 
registered bicycle rentals each vary by the prevailing temperatures. It also 
helps us appreciate that the relationship between bicycle rental numbers and
prevailing temperatures is non-linear and why temperature may not have too much 
of an influence on a linear model built to predict casual bicycle rentals.
</span>

```{r,  out.width = '50%', echo = FALSE, message = FALSE}
ggplot(aes(x=hum, y=casual), data=df_master) + 
  geom_bar(fill="#006633", stat="identity") + 
  xlab("Humidity") +
  ylab("Casual rental count") +
  ggtitle("Casual bicycle rentals Vs humidity")
ggplot(aes(x=hum, y=registered), data=df_master) + 
  geom_bar(fill="#006633", stat="identity") + 
  xlab("Humidity") +
  ylab("Registered rental count") +
  ggtitle("Registered bicycle rentals Vs humidity")
```

<span style="color:black">In the plot on the left, we see how casual bicycle
rentals stack up against the prevailing humidity. We see that normalized
humidity ranges between 0.30 and 0.55 seem to be very popular among casual
bicycle rentals.<br><br>In the plot on the right, we explore how registered
bicycle rentals stack up against the prevailing humidity. For registered
users, we see that normalized humidity values between 0.30 and 0.75 are most
popular among registered riders.<br><br>I chose to include these plots because 
it helps us appreciate that the relationship between bicycle rental numbers and
the prevailing humidity is non-linear and why the humidity factor is likely to
not have a great influence on any linear model we build to predict bicycle
rentals.</span>

```{r, echo=FALSE}
ggplot(aes(x=windspeed, y=cnt), data=df_master) + 
  geom_bar(fill="#006633", stat="identity") + 
  xlab("Windspeed") +
  ylab("Total rental count") +
  ggtitle("Total bicycle rentals Vs windspeed")
```

<span style="color:black">In the preceding plot, we see how total bicycle
rentals stack up against the prevailing wind conditions. We see that bicycle
rentals are highest when the windspeed is low, dropping off as the prevailing
windspeed picks up.<br><br>The correlation between bicycle rentals and windspeed 
is very low, as evidenced by the correlation chart we saw earlier. However, I 
chose to include this plot because it shows the distribution of bicycle rental 
numbers across the spectrum of windspeed and helps us appreciate the fact that 
there is no linear relationship between the two.</span>

```{r, echo=FALSE}
ggplot(aes(registered), data=df_master) + 
  geom_bar(fill="#006633", binwidth=5) + 
  ggtitle("Frequency plot of individual registered rentals")
```
<span style="color:black"><br>The preceding plot is a frequency plot of bicycle 
rentals. Most rentals happen in small numbers at a time, which shows that the
popularity of the service is still picking up. As popularity increases, we
should expect to see more occurences of higher number of bikes per day.</span>

<span style="color:black">Though the number of registered riders using the 
service is consistently higher than the number of casual renters, we see that 
the general trends remain the same. Registered users seem to be a little more 
loyal than casual users, in that the winter ridership is about the same as the 
summer ridership - whereas it is much lower in the case of casual users.<br><br>
I chose to include this plot in the report because this chart shows what we may
call the "popularity score" for the bicycle rental service as a whole.<br>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

We have already seen that average feeling temperatures and hour of the day are 
the main influencers of bicycle rentals. Let's look at how different weather 
conditions vary with the categorical variables.<br><br>We already have a feel 
for how feeling temperatures are connected with weather conditions. We will see 
how average feeling temperatures vary by season.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=atemp), data=df_master) + 
  geom_boxplot(fill="maroon") + 
  xlab("Prevailing season") +
  ylab("Normalized feeling temperatures") +
  ggtitle("Normalized feeling temperatures by season")
do.call(rbind, by(df$atemp, df$season, summary))
```

<span style="color:black"><br>The feeling temperatures are highest in fall and 
summer, followed by spring and winter. This follows the general trend of bicycle
rentals by season. I chose to plot this chart because I wanted to explore the
relationship between the different seasons and prevailing temperatures in a
given season. With a boxplot, I am able to see the median value for each season,
as well as the outliers.<br><br>Let's now look at the trend of humidity by 
seasons.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=hum), data=df_master) + 
  geom_boxplot(fill="maroon") + 
  xlab("Prevailing season") +
  ylab("Normalized humidity") +
  ggtitle("Humidity by season")
do.call(rbind, by(df$hum, df$season, summary))
```

<span style="color:black"><br>Humidity values are highest in winter, about the 
same in fall and summer, and lowest in spring. This is contrary to the trends we
see for bicycle rentals. This reinforces our finding that humidity is not a 
great predictor of bicycle rentals. I chose to include this plot because I 
wanted to explore the relationship between the humidity and season features. I
chose a boxplot because it lets us see the seasonal medians as well as any noisy
data points.<br><br>Let us take a look at how windspeeds influence bicycle 
rentals.</span>

```{r, echo=FALSE}
ggplot(aes(x=season, y=windspeed), data=df_master) + 
  geom_boxplot(fill="maroon") + 
  xlab("Prevailing season") +
  ylab("Windspeed") +
  ggtitle("Windspeed by season")
do.call(rbind, by(df$windspeed, df$season, summary))
```

<span style="color:black">Windspeeds are highest in spring and summer, and about
the same in fall and winter. This trend is contrary to the bicycle rental trends
.<br><br>From the preceding boxplots, we see that only average feeling 
temperature trends mimic that of bicycle rentals. The choice of the preceding
boxplots was driven by my desire to explore the relationship between features
and how they strengthen or weaken each other. I used boxplots because while a 
particular range of values of a given feature is dominant for the season,
unseasonal values do occur; I wanted to quantify how frequent these unseasonal
values would be.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

```{r, echo=FALSE}
reg <- lm(formula = cnt ~ registered, data = df_master)
summary(reg)
```

<span style="color:black"><br>The linear model above shows how the registered
user count influences the total count. We see that the correlation is pretty
high, with an r^2 value of 0.945.<br></span>

```{r, echo=FALSE}
reg <- lm(formula = cnt ~ temp, data = df_master)
summary(reg)
```

<span style="color:black"><br>The linear model above shows how the prevailing
temperatures influence the total bicycle rental count. There is a moderate
correlation, with an r^2 value of 0.1638.<br></span>

```{r, echo=FALSE}
reg <- lm(formula = cnt ~ hr, data = df_master)
summary(reg)
```

<span style="color:black"><br>The linear model above shows how the total bicycle
rentals are influenced by the hour of the day. We see that there is a moderate
correlation, with an r^2 value of 0.1552.<br></span>

```{r, echo=FALSE}
reg <- lm(formula = cnt ~ season, data = df_master)
summary(reg)
```

<span style="color:black"><br>The linear model above shows how the total bicycle
rentals are influenced by the prevailing season. We see that there is a slight
correlation, with an r^2 value of 0.06583.<br><br>I chose to include the linear
models above because, in anticipation of building a linear prediction model, I
am trying to decide on a set of features. From the correlation chart and other
plots above, we already have a fair idea of what features may have a strong
correlation with the bicycle rental numbers; the linear models above help us to
quantify that influence.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

# Bivariate analysis
<br>
<span style="color:black"><b><i>Talk about some of the relationships you 
observed in this part of the investigation. How did the feature(s) of interest 
vary with other features in the dataset?</i></b></span><br>
<span style="color:red">The number of bicycle rentals correlate strongly with 
the year and hour of the day. The former can be attributed to the assumption 
that the popularity of the rental service increased in the second year. This 
assumption is inconclusive because we only have data from two years. I do not 
see any particularly strong relationships between the number of bicycle rentals 
and other features in the dataset like humidity, weather temperature, feeling 
temperature, etc., (I consider those relationships that have an R^2 value > 0.50
to be a strong relationship). <br>From the exploration that I conducted into the
data, I see that about 16% of total bicycle rentals on a given day can be 
predicted by the prevailing temperature on that day.<br>I also see that a 
whopping 94% of total bicycle rentals on a given day can be predicted by the 
number of users that are registered with the service as on that day.<br>Given 
the preceding findings, I looked to see if there is a strong relationship 
between the number of casual bicycle renters and any other feature in the 
dataset. The strongest correlation I found was between casual ridership and 
normalized temperature - roughly 21% of casual bicycle rentals may be predicted 
by looking at the normalized prevailing temperatures for the day. </span>

<span style="color:black"><b><i>Did you observe any interesting relationships 
between the other features (not the main feature(s) of interest)?</i></b></span>
<br>
<span style="color:red">I looked at the relationships between some more features
in the dataset. I see that there is a moderate relationship between humidity and
the prevailing weather situation of the day. I see that there is a strong 
relationship between the normalized temperature on a given day and the 
corresponding season.</span><br>

<span style="color:black"><b><i>What was the strongest relationship you found?
</i></b></span><br>
<span style="color:red">The strongest relationship I found was between the total
bicycle rental numbers and number of registered renters. The other strong 
relationship I found was between the prevailing daily normalized temperatures 
and the season.</span><br>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

# Multivariate Plots Section
<br>
```{r, echo=FALSE}
ggplot(aes(x=DoW,y=cnt, group=factor(season), color=factor(season)),
       data=df_master)+ 
  stat_summary(fun.y=median, geom="line") + 
  xlab("Day of the week") +
  ylab("Total daily rental count") +
  ggtitle("Total daily bicycle rentals by prevailing season")
```

<span style="color:black">We see that daily median ridership for each day of the
week is lowest in spring and highest in the fall seasons. The daily medians 
numbers follow a closer trend early in the week in summer and winter. I chose
to include this plot because it shows very clearly the seasonal trends by day
of the week. From this plot, we see why the prevailing season may be an
acceptable indicator of bicycle rental numbers.</span>

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=season), data=df_master) + 
  scale_x_continuous(breaks=-1:24) + 
  stat_summary(fun.y=median, geom="line")  + 
  xlab("Hour of the day") +
  ylab("Total hourly rental count") +
  ggtitle("Total hourly bicycle rentals by prevailing season")
```

<span style="color:black">Looking at the median hourly ridership numbers, we see
that the numbers are lowest in spring, while they follow a closer trend in the 
other seasons. The numbers are pretty much the same in the early hours of the 
day(owing to the extremely low rentals). For hourly numbers in summer, fall and 
winter, we see they closely converge at around 9:00am and between 4:00pm and 
5:00pm - showing that these are primarily planned rentals. I chose to include
this chart because it supplements the investigation we performed in the previous
plot. When we look at the plot above, we see that the seasonal trends hold for
pretty much the entire range of hours.</span>

```{r, echo=FALSE}
ggplot(aes(x=DoW, y=cnt, color=weathersit, group=factor(weathersit)), 
       data=df) + 
  stat_summary(fun.y=median, geom="line") + 
  xlab("Day of the week") +
  ylab("Total daily rental count") +
  ggtitle("Total daily bicycle rentals by prevailing weather situation")
```

<span style="color:black">The preceding plot shows the median daily bicycle 
rentals depending on the prevailing weather situation of the day. Heavy 
precipitation seems to not deter bicycle renter on Mondays and Tuesdays, while 
on other days, heavy precipitation ensures low ridership. As predictable, 
highest bicyle ridership happens on clear days, followed by mild weather days 
and then on days that see light precipitation. I chose to include this plot
because it reveals the daily bicycle rental trends by prevailing weather
situation - and why there may not be a close correlation between the rental
numbers and prevailing weahter.</span>

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=weathersit), data=df) + 
  scale_x_continuous(breaks=-1:24) + 
  stat_summary(fun.y=median, geom="line") + 
  xlab("Hour of the day") +
  ylab("Total hourly rental count") +
  ggtitle("Total hourly bicycle rentals by prevailing weather situation")
```

<span style="color:black">Lastly, we look at median hourly bicycle rentals by 
weather situation. As predictable, heavy precipitation ensures lowest bicycle 
ridership except, curiously, in the earliest hours of the day when it is higher 
than in any other weather situation. Up until 9:00am, we see that ridership in 
clear weather and mild weather are almost the same. After 9:00am, rental numbers
in clear weather are higher than rental numbers in mild weather. Bicycle rentals
on days with light precipitation are higher than in heavy precipitation, but 
lower than in clear and mild weather, at all hours of the day. I chose to
include this plot because it nicely wraps up the investigation we have been
conducting over the past few plots. The trends are more pronounced for three of
the prevailing weather classes - and reveals why the correlation may be weak
between prevailing weather and bicycle rental numbers.</span><br><br>

<span style="color:black">These plots above show how the hourly and daily 
bicycle rentals depending on the weather situation and the seasons. I wonder how 
these trends would vary with temperature, humidity and windspeed factors too.
</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">
<br>
```{r, echo=FALSE}
ggplot(aes(x=season, y=cnt), data=df) + 
  geom_boxplot(fill="maroon") + 
  xlab("Prevailing season") +
  ylab("Total rental count") +
  ggtitle("Total bicycle rentals by prevailing season")
do.call(rbind, by(df$cnt, df$season, summary))
```

<span style="color:black"> The preceding plot shows the total bicycle rentals
cut by the season. Viewing the data as a boxplot shows us the median values, as
well as appreciate the fact that the rental numbers are more bunched up in
fall and summer(meaning that the total number of rentals stuck in a tight range)
whereas in spring and winter we saw a greater variance.
```{r, echo=FALSE}
ggplot(aes(x=weathersit, y=cnt), data=df) + 
  geom_boxplot(fill="maroon") + 
  xlab("Prevailing weather situation") +
  ylab("Total rental count") +
  ggtitle("Total bicycle rentals by prevailing weather situation")
do.call(rbind, by(df$cnt, df$weathersit, summary))
```

<span style="color:black">The preceding boxplots reinforce our findings that 
bicycle rentals are lowest in spring, followed by winter and summer and highest 
in the fall. Looking at the stats for each quartile, we see that the trends are 
close for the other three seasons except spring. We also see that periods of 
precipitation show lower bicycle rentals in the initial quartiles than for 
higher quartiles. <b><i>The main anomaly we see is that for the first quartile, 
bicycle rentals are higher in periods of heavy precipitation than in periods of 
light precipitation.</i></b>Towards the higher quartiles, though, we see that 
bicycle rentals in non-precipitative weather dominates.<br><br>I chose to 
include these plots because in addition to seeing the median numbers, we also 
get a feel for whether the bicycle rental numbers varied in a wide range or not.
<br>
<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

<br>Since bicycle rentals are dominated by registered users, and because we saw 
from prior analysis that bicycle rental trends closely match office/school-going
patters, I thought that it might be a good idea to explore the trends among 
casual users. I added a new column to the dataframe to represent the ratio of 
casual user rentals to the total renters.</span>

```{r, echo=FALSE}
df_master$casprop <- (df_master$casual + 1)/(df_master$cnt + 1)
df_master$season <- factor(df_master$season)
df_master$weathersit <- factor(df_master$weathersit)
ggplot(data=df_master, aes(casprop, fill=season)) + 
  geom_bar() + 
  ggtitle("Proportion of casual users to total rentals by prevailing season")
```

<span style="color:black">We see that the bulk of ratio frequencies is below 1:2
casual:total rentals. Of these, we see that casual ridership is highest in 
winter, followed by summer, spring and fall. I think this may be because winter 
ridership may be primarily ad-hoc, while in better weather, people make plans 
and register for bicycle rentals. I chose to include this plot because it 
reinforces our earlier findings that bicycle rentals are dominated by registered
users. Coloring the numbers by the prevailing season, we easily see the variance
by season.</span>

```{r, echo=FALSE}
ggplot(data=df_master, aes(casprop, fill=weathersit)) + 
  geom_bar() + 
  ggtitle("Proportion of casual users to total rentals by prevailing weather")
```

<span style="color:black">The highest occurences of casual ridership is seen in 
mild and light precipitative weather, followed by clear weather. There is very 
little casual ridership in heavy precipitation. This may be because casual 
riders are those who rent bicycles ad-hoc in slightly inclement weather, while 
they make plans and register for the service when clear weather is expected. I 
chose to include this plot because it reinforces our earlier findings that 
bicycle rentals are dominated by registered users. Coloring the numbers by the 
prevailing season, we easily see the variance by the prevailing weather
situation.</span>

```{r, echo=FALSE}
ggplot(data=df_master, aes(casprop, fill=DoW)) + 
  geom_density(alpha=0.5) + 
  ggtitle("Proportion of casual users to total rentals by day of the week")
```

<span style="color:black">On plotting the frequency of casual ridership ratio to
the total by day of the week, we see that casual ridership is higher earlier in 
the week than later. Interestingly, we see that casual ridership is higher on 
Sundays than on Saturday. I chose to include this plot because it reinforces our
earlier findings that bicycle rentals are dominated by registered users. 
Coloring the numbers by the prevailing season, we easily see the variance by the
prevailing weather situation.</span>

```{r, echo=FALSE}
ggplot(data=df_master, aes(casprop, fill=as.factor(holiday))) + 
  geom_density(alpha=0.5) + 
  ggtitle("Proportion of casual users to total rentals by holiday/working day")
```

<span style="color:black">Finally, on plotting the frequencies of casual 
ridership ratios to the total by the fact that the days are holidays/working 
days, we see that casual ridership is higher on working days than on holidays. 
We see that on holidays, we see higher ratios of casual riders to total rentals.
This may be explained by the fact that people like to explore the city on 
holidays.<br><br>I chose to include this plot because, in prior charts and 
analyses, we concluded that registered rentals dominate total bicycle rental 
numbers. That leaves us wondering as to whether this trend holds on 
"non-commuter days", ie., holidays. This chart explains that question and clears 
the air.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

<span style="color:black">All of these plots reinforce the findings that we 
reached using earlier histograms and plots. Now, in order to build a linaear 
model to predict bicycle ridership, I'm going to decide on a set of variables 
that correlate with ridership.</span>


```{r, echo=FALSE}
ggplot(df_master, aes(hr, cnt), fill=hr) + 
  geom_boxplot(fill="maroon") + 
  scale_y_log10() + 
  scale_x_continuous(breaks=-1:24) +
  facet_wrap(~ season) +
  xlab("Hour of the day") +
  ylab("Total hourly rental count") +
  ggtitle("Hourly bicycle rentals by the prevailing season")
```

<span style="color:black"><br>I chose to include this plot because we already
found out, from previous charts and analyses, how certain hours of the day are
popular with bicycle renters. I wanted to see how the bicycle rental numbers 
varied by hour of the day and the prevailing seasons, and how the data points 
were distributed. We see that the trends in this chart follow the trends we saw
earlier, that the middle portion of the day is a clear favorite among bicycle 
renters except in spring, where the trends are a little diluted.</span>

```{r, echo=FALSE}
ggplot(df_master, aes(hr, cnt), fill=hr) + 
  geom_boxplot(fill="maroon") + 
  scale_y_log10() + 
  scale_x_continuous(breaks=-1:24) +
  facet_wrap(~ weathersit) +
  xlab("Hour of the day") +
  ylab("Total hourly rental count") +
  ggtitle("Hourly bicycle rentals by prevailing weather")
```

<span style="color:black">From the plots above, we see how clear weather is a
favorite among bicycle renters, with higher number of hourly rentals happening
in this weather. The next favorite is mild weather. Periods of light 
precipitation see moderate rentals, while periods of heavy precipitation see
almost no rentals. I chose to include this plot because we already found out, 
from previous charts and analyses, how certain hours of the day are popular with
bicycle renters and wanted to see how the bicycle rental numbers varied by hour 
of the day and the prevailing seasons, and how the data points were distributed.
<br><br>From the plots above and the r^2 values of the linear models above, we 
can deduce that the features that may help us predict total bicycle rentals on 
any given day are "registered", "temp", "hr", "season", "weathersit", in the 
order of confidence. Let's try to build a linear prediction model using these 
features and see if we can predict bicycle rental numbers with a high level of 
confidence.</span>

```{r, echo=FALSE}
m1 <- lm(I(cnt) ~ I(registered), data=df_master)
m2 <- update(m1, ~ . + temp)
m3 <- update(m2, ~ . + hr)
m4 <- update(m3, ~ . + season)
m5 <- update(m4, ~ . + weathersit)
m6 <- update(m5, ~ . + windspeed)
mtable(m1, m2, m3, m4, m5, m6)
```

<span style="color:black"><br>To the variables I mention in the section above, I 
added the "windspeed" variable to see if factoring it in would give my 
prediction model some more accuracy. We see that the prediction model I just 
built can help predict bicycle rentals on a given day with about 95.5% 
confidence.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

# Multivariate Analysis

<span style="color:black"><b><i>Talk about some of the relationships you 
observed in this part of the investigation. Were there features that 
strengthened each other in terms of looking at your feature(s) of interest?</i>
</b></span><br>
<span style="color:red">Many of the variables interact and strengthen each 
others' influence over bicycle rental numbers. For example, temperature, 
humidity and windspeed are tightly related to the weather situation. Each season
has a weather situation that dominates a majority of the days in that season.
</span>

<span style="color:black"><b><i>Were there any interesting or surprising 
interactions between features?</i></b></span><br>
<span style="color:red">I did not see any surprising relationships between 
features in the dataset.</span>

<span style="color:black"><b><i>Did you create any models with your dataset? 
Discuss the strengths and limitations of your model.</i></b></span><br>
<span style="color:red">Yes, I created a prediction model with the dataset. 
The main strength of my prediction model is that it is able to predict bicycle 
rental numbers with about 95.5% confidence. Some limitations of the model: 
<br>1) The dataset spans only about two years. As we collect more and more data,
we may need to tweak the model.<br>2) The model does not take into account any 
special events happening in the city on a given day(this is actually a 
limitation of the dataset!). If we know whether there were any significant 
events in the city(rally, game, etc.) on a given day, we could have used that as
an input into our model.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

# Final Plots and Summary

## Plot One

```{r, echo=FALSE}
ggplot(hr_casreg, aes(hr, value, col=variable)) + 
  scale_x_continuous(breaks=-1:24) + 
  stat_summary(fun.y=median, geom="line") + 
  xlab("Hour of the day") +
  ylab("Median hourly rental count") +
  ggtitle("Median hourly bicycle rentals")
```

## Description One

<span style="color:black">In this plot, we see how the registered bicycle rental
numbers trump casual bicycle rentals. Looking at these trends gave us an early
indication that looking at the number of registered rentals per hour/day would
give us a good idea of what the total bicycle rental numbers would look like.
</span>

## Plot Two

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=season), data=df_master) + 
  scale_x_continuous(breaks=-1:24) + 
  stat_summary(fun.y=median, geom="line") + 
  xlab("Hour of the day") +
  ylab("Median hourly rental count") +
  ggtitle("Total hourly bicycle rentals by prevailing season")
```

## Description Two

<span style="color:black">In the previous plot, we see the hourly trends for 
bicycle rentals across all the observations in the dataset, colored by seasons. 
In all seasons, we see that one of the hours peaks occurs at 8:00 AM and the 
second one at 5:00 PM. These peaks suggest that bicycle rentals are utilized by 
users who use the bikes to get to work/school(since these activities tend to be 
between fixed times everyday). We see that the cicycle rental numbers fall on 
either side of the peak, marking their lowest points during the night hours. The
general trend holds in all seasons, but the numbers are highest in the fall, 
followed by summer, winter and spring. I think this is mainly because of 
temperature, precipitation and other factors associated with the seasons.</span>

## Plot Three

```{r, echo=FALSE}
ggplot(aes(x=hr, y=cnt, color=weathersit), data=df_master) + 
  scale_x_continuous(breaks=-1:24) + 
  stat_summary(fun.y=median, geom="line")  + 
  xlab("Hour of the day") +
  ylab("Total hourly rental count") +
  ggtitle("Median hourly bicycle rentals by prevailing weather situation")
```

## Description Three

<span style="color:black">In the previous plot, we see hourly trends for bicycle
rentals across all observations in the dataset, colored by the prevailing 
weather situation. The trends we see in the previous plot are, to some extent, 
seen in this plot too. Let's consider the three weather situations - clear, mild
and light precipitation first. We see the conventional "commuter peaks" at 8:00 
AM and 5:00PM, with ridership falling off on either sides of the peaks. 
Interestingly, renter behavior is very different during periods of heavy 
precipitation. Total rentals during heavy precipitation stays close to none in 
the early hours of the day, peaking slightly at around 6:00 PM.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">

# Reflection

<span style="color:black">The bicycle rentals dataset consists of 17379 hourly 
observations of bicycle rentals at a rental agency in Washington DC. I chose 
this dataset because I wanted to gain some experience in choosing a tidy dataset
to explore. I see that it is mostly weather-related features that closely 
influence the bicycle rental numbers on a given day. Factors like temperature, 
precipitation and windspeed tie closely into the prevailing weather situations 
and seasons. Therefore, to some extent, these factors influence and strengthen 
each other. Another feature that influences bicycle rentals is the hour of the 
day. I think this may be driven by office/school-goers who use bicycles to 
commute - and I see peaks occuring at the conventional working hours. Also, as 
expected, we see lesser bicycle rentals during the cold season and during 
periods of heavy precipitation.<br><br>I searched for a long time and at various
sources to find a good, tidy dataset to practice my skills on. Many datasets 
were interesting, but did not meet the conditions for this project. I downloaded 
and melted several datasets from the gapminder collection and explored them, 
only to find that the main features did not influence each other and the 
prediction model that I tried to put together was very weak. Looking back, I 
consider those attempts to be good ones to cut my teeth on, just that they 
weren't impressive enough for a final project.<br><br>Ultimately, I found the 
bicycle rentals dataset, which met all the criteria for the project. One of the 
first things I did was to plot a ggpairs chart and put together a linear 
regression model. When I had concluded that the dataset was interesting enough, 
I set out to explore it in a systematic manner. During the course of the 
project, I was able to put together a linear prediction model for bicycle 
rentals. The model is able to predict bicycle rentals with a 95% confidence 
level.<br><br>In conclusion, I would like to correlate this dataset with events 
that happened in Washington DC. on each of these days. That would help me see if 
casual bicycle rentals were driven by people wanting to avoid event-driven 
crowds on the popular transportation systems. Also, I would love to explore the 
data for a few more years than contained in this dataset.</span>

<hr style="BORDER-RIGHT: medium none; BORDER-TOP: #cccccc 2px solid; 
BORDER-LEFT: medium none; BORDER-BOTTOM: medium none; HEIGHT: 1px">